Beyond the best guess: Distributional predictions from large language models for human forced-choice data

### 2023-10-06 03-paper/paper-LLM-probmod.tex
rearranged and revised §5 & 6

To keep matters simple, we use an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks, so-called \textbf{reference games} \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
In the \textbf{production condition}, the speaker is assigned a \emph{target object} from the context set which they have to describe to the interpreter.
In the \textbf{interpretation condition}, the interpreter observes a description, here called \emph{trigger word}, and chooses one of the objects from the context set.
The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the target object; and, for the interpreter, to guess correctly which object the speaker had in mind.

@@

The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts} on the left.\footnote{
  The production condition actually has two distractor choices.
  Here and in the following, these are lumped together as a single category, also when modeling random errors in later models.
  This is a simplification but does not change anything of substance.}
It is interesting that the distractor options were chosen rather often.
We also see that the number of target choices is higher in the production condition than in the interpretation condition.
This is in line with previous experimental results on human reference games.
% \citep{QingFranke2013:Variations-on-a}.
For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).

@@

    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of three models (shapes and error bars).
      Shapes show the mean of the posterior predictive distributions of three models, differing in the scope of the item-level aggregation with respect to soft-maximizing and scaling.
      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
    }
  \label{fig:refgame-counts}
\end{figure}

@@

Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,FrankeDegen2015:Reasoning-in-Re} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.

@@

\begin{figure}
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model (with $\alpha=1$).
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents), assume $\alpha=1$ and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

Figure~\ref{fig:RSA-example} gives example calculations (assuming flat priors and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
For $\alpha=1$, the model predicts that the probability of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
Increasing $\alpha$ will increase the odds of target over competitor choices.
Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figur~\ref{fig:refgame-counts}e, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a random error probability for each choice \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, such as via Laplace smoothing: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
\begin{align*}
  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) \propto P_{r}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}\,,
\end{align*}
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{LLM predictions for reference games}
\label{llm-predictions-for-reference-games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section explores different ways of deriving a probabilistic likelihood function for the human data from a Large Language Model.
For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
% The previous literature has shown that scoring longer sequences of text for an autoregressive LLM in terms of \emph{average next-token log probabilities}, or \emph{average log-probs} for short, gives improved predictions in relevant benchmark tests \citep[e.g.,][]{BrownMann2020:Language-Models}.
Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
In the following, we first describe the different options of deriving LLM-based predictor terms.
We then compare all probabilistic models based on their adequacy of explaining the human choice data.

@@

\subsection{LLM-based probabilistic predictions for forced-choice data}
\label{sec:notat--term}

Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
}
Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
We will define an item-level (non-normalized) score $S(R_{l}, I_{k})$ for each response type $R_{l}$ and item $I_{k}$, and use this to define a condition-level probabilistic predictions for response types by averaging over all items in a given condition.

\paragraph{Item-level score.}
The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $k$ is defined as the log-probability of continuation $O_{ki}$ after prompt $I_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
%
\begin{align*}
S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.

\paragraph{Condition-level predictions.}
Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  % & \textbf{narrow-scope aggregation} \\
  & P_{n}(R_{l}, C ; \alpha) \propto \expo \left [  \alpha \ \frac{1}{m} \ \sum_{i = k}^{m} S(R_{l}, I_{k})  \right ]
    \tag*{\textcolor{gray}{[narrow-scope aggregation]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{w}(R_{l}, C ; \alpha) \propto \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( \alpha \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha \ S(R_{l'}, I_{k}) \right )}
    \tag*{\textcolor{gray}{[wide-scope aggregation]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{i}(R_{l}, C ; \alpha) \propto  \left [ \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( S(R_{l'}, I_{k}) \right )} \right ]^{\alpha}
    \tag*{\textcolor{gray}{[intermediate-scope aggregation]}}
\end{align*}

All three condition-level predictors coincide when there is only one item, of course.
For more items, wide-scope and intermediate-scope aggregation coincide when $\alpha = 1$.
In all other cases, the predictors are not guaranteed to be identical.
Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
Narrow-scope aggregation considers raw scores, which are \emph{not} probabilistic predictions at the item level.
Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark test for $\alpha \rightarrow \infty$.
Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.

\subsection{Model fitting, criticims \& comparison}
\label{sec:model-fitting}

Parameterized predictions $P_{m}(R_{l}, C ; \alpha_{c})$ for models $m \in \set{n,w,i,r}$, where $r$ is the RSA model introduced in Section~\ref{sec:model-pred-from}, can be assessed in the light of the empirical data from human participants with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
For each condition $C$ (production and interpretation), each model has a free optimality parameter $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ with a log-Normal prior.
Following the rationale from Section~\ref{llm-predictions-for-reference-games}, we also assume that each model has one epsilon parameter per condition, $\epsilon_{c}$, which accounts for random errors in the human data via Laplace smoothing, with a prior $\epsilon_{c} \sim \text{Beta}(1,5)$ favoring small values.
The resulting likelihood function for model $m$ and data from condition $C$ is a multinomial distribution with predictor for response $R_{l}$ given by:
%
\begin{align*}
  P_{m}(R_{l}, C; \alpha_{c}, \epsilon_{c}) \propto P_{m}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}
\end{align*}

Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of each $\alpha_{c}$ and $\epsilon_{c}$ for each model.
The usual Bayesian summary statistics for posteriors over model parameters are shown in Figure~\ref{fig:posterior-stats}.

@@

  \caption{
    Summary statistics of posterior samples.
    Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
    \mf{maybe not necessary, but can we explain the high values for $\alpha$ in the wide model?}
  }
  \label{fig:posterior-stats}
\end{table}

To assess goodness-of-fit, Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution, i.e., the model's predictions about data of the same size and structure as the training data.
As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
Figure~\ref{fig:refgame-counts} shows that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
Consequently, the results from Table~\ref{tab:Bppp-values} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Item-level predictions}
\label{sec:item-level-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The ability to predict a condition-level average is useful for applications, and may also be theoretically insightful.
Nevertheless, for human data, averages over item- or individual-level variation can also be misleading \citep[e.g.,][]{StanovichWest2000:Individual-diff,EstesMaddox2005:Risks-of-Drawin,HaafRouder2019:Some-do-and-som}, as aggregate patterns may be artifacts of systematic variation at the subject- or item-level.
LLMs do not make individual-level predictions, at least not without further intervention, such as special prompting.
But the question remains whether the variation in item-level predictions made by LLMs is reflected in the human data.

>>>>

Reference games are an established, well-understood and austere experimental paradigm to test human decision making in abstract communicative tasks \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag,DegenFranke2013:Cost-Based-Prag,QingFranke2013:Variations-on-a,Frank2016:Rational-speech,SikosVenhuizen2021:Reevaluating-pr}.
A reference game consists of two players, a speaker and an interpreter, who jointly observe a set of objects, usually referred to as context (see Figure~\ref{fig:ref-game}).
In the \textbf{production condition}, the speaker is assigned a \emph{target object} from the context set which they have to describe to the interpreter.
In the \textbf{interpretation condition}, the interpreter observes a description, here called \emph{trigger word}, and chooses one of the objects from the context set.
The goal of the game is, for the speaker, to choose a description that enables the interpreter to choose the target object; and, for the interpreter, to guess correctly which object the speaker had in mind.

@@

The overall distribution of choices that correspond to the target, competitor, and distractor states is shown in Figure~\ref{fig:refgame-counts}.\footnote{
  The production condition actually has two distractor choices.
  Here and in the following, these are lumped together as a single category, also when modeling random errors in later models.
  This is a simplification but does not change anything of substance.}
It is interesting that the distractor options were chosen rather often.
We also see that the number of target choices is higher in the production condition than in the interpretation condition.
This is in line with previous experimental results on human reference games.
% \citep{QingFranke2013:Variations-on-a}.
For example, in previous forced-choice reference games with human participants with pictorial presentations of objects, \citet{QingFranke2013:Variations-on-a} observed the following proportions: $\tuple{0.882, 0.118, 0}$ in the production and $\tuple{0.492, 0.506, 0.003}$ in the interpretation condition (for 288 observations in each condition).

@@

    \caption{Counts of choices from reference games with human participants (colored bars), with summary statistics from the posterior predictive distribution of four models (shapes and error bars).
      Shapes show the mean of the posterior predictive distributions of the RSA model and three further models derived from item-level LLM-supplied scores.
      Error-bars show corresponding 95\% credible intervals of the posterior predictive.
    }
  \label{fig:refgame-counts}
\end{figure}

@@

Data from reference games with human participants have been variously analyzed with probabilistic models using inspiration from behavioral game theory \citep[e.g.,][]{DegenFranke2013:Cost-Based-Prag}, probabilistic Bayesian modeling \citep[e.g.,][]{FrankGoodman2012:Predicting-Prag} or other forms of probabilistic modeling \citep[e.g.,][]{GattGompel2013:Are-we-Bayesian}.
Common to these approaches is that they derive or define, based on some explicit conceptual motivation, a parameterized stochastic speaker policy, $P_{S}(u \mid s; \theta_{S})$, modulated by parameters $\theta_{S}$, for a speaker's choice of expression or utterance $u$ given a referent or state $s$, which the speaker wants to communicate;
and a stochastic listener policy, $P_{L}(s \mid u; \theta_{L})$, capturing the probability of choosing a referent $s$ for utterance $u$.

@@

\begin{figure}[t]
  \centering
  \includegraphics[width = 0.9 \textwidth]{00-pics/RSA-example.png}
  \caption{
    Example of predictions from the RSA model (with $\alpha=1$).
    The semantic meaning is shown as a matrix of binary truth-values.
    The policies of literal listener, pragmatic speaker and listener are calculated for uniform priors over state (referents), assume $\alpha=1$ and are shown as row-stochastic matrices.
  }
  \label{fig:RSA-example}
\end{figure}

Figure~\ref{fig:RSA-example} gives example calculations (assuming flat priors and $\alpha=1$) for the reference game from Figure~\ref{fig:ref-game}.
For $\alpha=1$, the model predicts that the probabilities of target, competitor and distractor options are $\tuple{\nicefrac{2}{3}, \nicefrac{1}{3}, 0}$ for the production, and $\tuple{0.6, 0.4, 0}$ for the interpretation condition.
Increasing $\alpha$ will increase the odds of target over competitor choices.
Yet, the model also predicts probability zero for distractor choices, so that the human data shown in Figure~\ref{fig:refgame-counts}, where the distractor option was chosen in both conditions, would immediately rule out the model entirely.
It is therefore common to include a random error probability for each choice \citep[e.g.,][]{LeeWagenmakers2013:Bayesian-Cognit}, such as via Laplace smoothing: if $P_{r}(R_{l}, C, \alpha_{c})$ is the RSA model's probabilistic prediction for response category $R_{l}$ (target, competitor, or distractor) for condition $C$ (production or interpretation) and condition-specific optimality $\alpha_{c}$, the random-error smoothed prediction is:
\begin{align*}
  P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c}) \propto P_{r}(R_{l}, C; \alpha_{c}) +  \frac{\epsilon_{c}}{3}\,,
\end{align*}
where $\epsilon_{c}$ is a (condition-specific) parameter giving the probability that a choice was made by randomly guessing.
The result is a four-parameter model, one pair of parameters per condition, which provides a likelihood function for the categorical choice data and so can be fitted to the data and compared against other probabilistic models providing a likelihood function for the same data.

Parameterized predictions like $P_{r}(R_{l}, , C; \alpha_{c}, \epsilon_{c})$ can be assessed in the light of the empirical data with the usual tools of Bayesian data analysis \citep[e.g.][]{GelmanCarlin2014:Bayesian-Data-A,McElreath2016:Statistical-Ret,Lambert2018:A-Students-Guid}.
Let $\alpha_{c}\sim \text{log-Normal}(0.5,1)$ have a reasonably wide log-Normal prior, and let $\epsilon_{c} \sim \text{Beta}(1,5)$ have a Beta prior favoring small values.
Using Stan \citep{Team2023:The-Stan-Core-L} for Bayesian inference, we obtain estimates of posterior credible values of model parameters (summary statistics of which are shown in Table~\ref{fig:posterior-stats}).

To assess goodness-of-fit, we can use the posterior predictive distribution, i.e., the model's predictions about data of the same size and structure as the training data.
As a minimal bar, we would require a model trained on observed data $D_{\text{obs}}$ to not be surprised by $D_{\text{obs}}$.
Figure~\ref{fig:refgame-counts} shows summary statistics (means and 95\% credible intervals) for the posterior predictive distribution of the RSA model (among other models to be introduced later).
We see that the RSA model passes this ``visual posterior predictive check'' \citep{Kruschke2021:Bayesian-Analys} for both conditions.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
These values approximate the probability that a model trained on $D_{\text{obs}}$ would predict future data of the same size and format that is at least as unlikely as the data $D_{\text{obs}}$ itself.
Low values therefore indicate that the trained model would be highly surprised by the data it was trained on; an indication that it failed to capture something essential about the training data.
The results from Table~\ref{tab:Bppp-values} show that the RSA model passes this numerical test of model adequacy.

The following sections apply these test of Bayesian model criticism also to models built around predictor values from LLMs.
Section~\ref{sec:item-level-pred} first looks at the data from each item individually, before Section~\ref{llm-predictions-for-reference-games} investigates different ways for deriving condition-level predictions.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Item-level predictions from LLMs}
\label{sec:item-level-pred}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Figure~\ref{fig:refgame-counts} shows the condition-level counts for choices for all of the 100 different items combined.
An LLM, on the other hand, first and foremost makes predictions about each item.
The items of the reference game experiment differ in the task-relevant features (color, shape, texture), as well as the order of presentation of states and words.
We expect both human and LLM predictions to vary between different items: e.g., humans seem to have preferences for some features \citep{QingFranke2013:Variations-on-a}, maybe even leading to over-production of informationally irrelvant material \citep{DegenHawkins2020:When-redundancy}; and machines may be susceptible to the presentation of the order of choice options. \mf{references from the NLP literature on effects of prompt variation?}
The main question to be addressed in this section is whether a standard item-level score from LLMs provides a good fit, if we aim at predicting the human data separately for each item, not as a condition-level average.

For a set-up in which the task descriptions and choice options are all text-based, predictions of LLMs at the item-level can be derived as a function of the (log) probabilities assigned to the continuations corresponding to each choice option after processing the task description.
Let $C$ be the experimental condition of interest (here: the production or interpretation condition).
Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
}
Each item $I_{k}$ consists of a task description $T_{k}$ and $l$ choice options $O_{k1}, \dots, O_{kl}$, which are treated as textual continuations of the task description (see Appendix~\ref{sec:examples-items-llm} for an example).
For each item $I_{k}$, let $F(R_{l}, I_{k})$ be the choice option $O_{ki}$ that corresponds to \emph{response type} $R_{l}$ (here: target, competitor, distractor).\footnote{In the current set-up the response type ``distractor'' has two instantiations in the production condition. Since choice options are a single word in the production condition, for simplicity, we lump both of the distractor words together and treat them as a single option by taking the sum of the log-probabilities for both distractor words.}
The \emph{item-level score} $S(O_{{ki}}, I_{k})$ for option $O_{ki} = w_{ki1}, \dots w_{kin}$ of item $I_{k}$ is defined as the log-probability of continuation $O_{ki}$ after prompt $T_{k}$:\footnote{Numerically, we worked with \emph{average} log-probabilities. As in our experiment, all choice options had the same number of words, this is largely irrelevant, except for the effect of the $\alpha$ parameter in the interpretation condition. Using length-corrected log-probabilities is the same as using raw log-probabilities with an $\alpha$ parameter divided by the number of words. In this way, we make the $\alpha$ parameter more meaningfully comparable across conditions.}
%
\begin{align*}
S(O_{ki}, I_{k}) =  \sum_{j=1}^n \log P_{\text{LLM}} \left(w_{kij} \mid T_{k}, w_{ki1}, \dots, w_{ki(j-1)} \right)  \,.
\end{align*}
%
The item-level score of response type $R_{l}$ is $S(R_{l}, I_{k}) = S(F(R_{l}, I_{k}))$.

Using item-level scores from GPT-3.5, we fit a model to the human data at the item-level.
The data to be explained is the set of all triples of response type counts $\tuple{t_{k}, c_{k}, d_{k}}$ (target, competitor, distractor), separately for each item $I_{k}$.
The model's predictions for the data from item $I_{k}$ from condition $C$ (production or interpretation) are:
%
\begin{align*}
  P_{\text{item}}(R_{l}, C, I_{k}, \alpha_{c}, \epsilon_{c}) \propto \frac{\expo \left( \alpha_{c} \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha_{c} \ S(R_{l'}, I_{k}) \right )} + \epsilon_{c}\,.
\end{align*}
%
In words, we use a single pair of parameters $\alpha_{c}$ and $\epsilon_{c}$, one for each condition, take the LLM's item-level scores for each item, and derive the response probabilities in the exact same way as before.
% Notice that item-level scores are the same for all three types of models introduced and compared in the previous section, so that this by-item analysis applies to the raw LLM prediction irrespective of down-stream aggregation.
Priors on parameters are as for the RSA model described in Section~\ref{sec:model-pred-from}.

Figure~\ref{fig:item-level-obs-pred} shows that the item-level LLM scores predict variance which is not borne out by the human data.
Concretely, the plot shows, for each item, mean posterior estimates of the model's predicted probability of choosing the target option ($x$-axis), together with the observed proportion of target choices in the human data ($y$-axis).
There is ample variation in the model's predictions, especially visible in the production condition, owing to the fact that the item-level scores of the LLM sometimes clearly favor another option than the target choice.
So, the model itself predicts systematic variability at the item level.
The human data, too, show variability at the item-level, but there is no (visual) indication that the item-level variability predicted by the LLMs is borne out by the human data.

Sampling-based approximations of Bayesian posterior predictive $p$-values for the by-item analysis are very low (0.0001 for production, 0 for interpretation), suggesting that the unaggregated item-level LLM predictions are inadequate predictors of the human data.
In contrast, using the global, item-independent RSA model predictions to fit the item-level data, we obtain estimates of posterior predictive $p$-values that do not discredit the model ($\sim 0.338$ for production, $\sim 0.252$ for interpretation).
These results suggest that LLM-based probabilistic predictions may imply item-level variance that is not attested in the human data.
Put more strongly, a model that seeks to predict what human participants choose on a by-item level, based on the most obvious item-level score, is ruled out by experimental data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Condition-level predictions}
\label{llm-predictions-for-reference-games}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

This section explores different ways of deriving a probabilistic likelihood function for the human data from a Large Language Model at the aggregate, condition-level.
Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
In the following, we first describe the different options of deriving LLM-based predictor terms.
We then compare all probabilistic models based on their adequacy of explaining the human choice data.

@@

Condition-level predictions are a probability distribution over response types, obtained from item-level scores $S(R_{l}, I_{k})$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  % & \textbf{narrow-scope aggregation} \\
  & P_{n}(R_{l}, C ; \alpha) \propto \expo \left [  \alpha \ \frac{1}{m} \ \sum_{i = k}^{m} S(R_{l}, I_{k})  \right ]
    \tag*{\textcolor{gray}{[narrow-scope aggregation]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{w}(R_{l}, C ; \alpha) \propto \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( \alpha \ S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( \alpha \ S(R_{l'}, I_{k}) \right )}
    \tag*{\textcolor{gray}{[wide-scope aggregation]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{i}(R_{l}, C ; \alpha) \propto  \left [ \frac{1}{m} \ \sum_{i = k}^{m} \frac{\expo \left( S(R_{l}, I_{k}) \right )}{ \sum_{l'} \expo \left( S(R_{l'}, I_{k}) \right )} \right ]^{\alpha}
    \tag*{\textcolor{gray}{[intermediate-scope aggregation]}}
\end{align*}

All three condition-level predictors coincide when there is only one item, of course.
For more items, wide-scope and intermediate-scope aggregation coincide when $\alpha = 1$.
In all other cases, the predictors are not guaranteed to be identical.
Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
Narrow-scope aggregation considers raw scores, which are \emph{not} probabilistic predictions at the item level.
Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark tests for $\alpha \rightarrow \infty$.
Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.

\begin{table}[t]
  \centering

@@

  \caption{
    Summary statistics of posterior samples.
    Number indicate estimates for lower and upper bounds of 95\% credible intervals and the posterior means.
  }
  \label{fig:posterior-stats}
\end{table}

The three condition-level predictions yield three different probabilistic models, each with two pairs of condition-dependent parameters $\alpha_{c}$, and $\epsilon_{c}$.
Using priors and Bayesian posterior estimation as previously for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and posterior predictive distributions.
The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{fig:posterior-stats}.
Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range in both conditions.
This is because, already for values of $\alpha_{c} = 1$, the item-level predictions are virtually ``winner-takes-all'' choices, selecting the option with the maximum score with probability almost 1 for almost all items.
This implies that, by studying the posterior predictive distribution of the wide-scope model, we are virtually testing condition-level predictions from the standard WTA-strategy used for benchmark testing (see Section~\ref{motivation}).

Figure~\ref{fig:refgame-counts} shows that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate.
To corroborate the visual impression, Table~\ref{tab:Bppp-values} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:Bppp-values} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.

@@
